\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Update local trees section

% % First authors
% \author[1]{Yan Wong}

% % Second authors
% \author[2,$\star$]{Anastasia Ignatieva}
% \author[3,4,$\star$]{Jere Koskela}

% % Middle Authors
% \author[5]{Gregor Gorjanc}
% \author[6,7]{Anthony W. Wohns}

% % Corresponding
% \author[1,$\dagger$]{Jerome Kelleher}

% \affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
% \affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
% \affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
% \affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
% \affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
% \affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
% \affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}

% \affil[$\star$]{Denotes shared second authorship, listed alphabetically}
% \affil[$\dagger$]{Denotes corresponding author}

@@

% FIXME abstract needs some rewriting, we're placing too much emphasis
% on the coalescent there.
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on specific details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
We show that eARGs are limited in the patterns
of genetic inheritance that can be represented,
and require a precise description of the details of all recombination events.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[][AND OTHERS]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is a lack of consensus on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress in applying genetic genealogies to understand the biological world.

@@

It is useful to have
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% This sentence needs another pass
A node in an oriented tree can have
zero, one (see the next section for discussion of ``unary'' nodes),
or many children (a node with more than two children is known as a polytomy).
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree (or in any other form in which
the tree nodes are explicitly labelled with the identity of the
corresponding ARG node)
along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

The internal nodes of trees, however, are usually
considered to be \emph{unlabelled},
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
using only the tree topologies it would not be straightforward to
infer that node \noderef{i} in the first and last trees of
Fig.~\ref{fig-arg-in-pedigree}C is the same node.
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
Their statement that an ARG
contains more information than its local trees is true only if
the local trees are unlabeled and/or do not contain unary nodes
(see the next section for discussion on this point).
However, if the local trees
contain labelled internal nodes,
then there is precisely as
much information in them as in an ARG.

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
sequentially along the genome, and subsequently stitching them together
into an ARG.
Two recent methods take this approach,
\relate~\citep{speidel2019method} and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among a small number of samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

% The equivalence between an ARG and a set of local trees
% separated by SPRs is often mentioned [e.g. x y z], but
% it is important to note that while extracting local
% trees from an ARG is straightforward, the converse
% process of reconstructing precisely the same ARG from
% these trees is more problematic. If the internal
% nodes are consistently labelled and unary nodes are
% included in the local trees (e.g., Fig~\ref{fig-simplification})
% then it is easy to exactly reconstruct the ARG.
% Otherwise, the problem is much more difficult and it
% may not be possible to uniquely reconstruct the original ARG.
% Without labels for the internal nodes, we are forced to
% to reason about which subtrees remain the same and which
% differ due to the SPR separating adjacent trees in order to
% assign node identity. Without the internal unary nodes marking
% recombinations, we must infer their existence. In
% general, there will not be a unique solution to this problem,
% and so there can be many possible ARGs compatible with
% a given list of leaf-labelled binary trees.
% This is the approach taken by the Espalier inference
% method~\citep{rasmussen2022espalier}, which finds a
% set of SPRs that reconcile independently inferred
% trees from different regions of the genome.

@@

TODO mention \citep{schaefer2021ancestral} somewhere.
Also \citep{kuhner2017consensus}.

@@

An alternative approach is to treat the ARG as a latent parameter to be
averaged out by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian,
fallon2013acg, mahmoudi2022bayesian}. These methods operate on representations
of the ``little ARG'', and are extremely computationally expensive, being
applicable to at most hundreds of samples consisting of tens or hundreds of
kilobases with human-like parameters. State-of-the-art methods rely on cheaper,
approximate models \citep{didelot2010inference, heine2018bridging,
hubisz2020mapping,hubisz2020inference, medina2020speeding}. The most scalable
method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to
dozens of mammal-like genomes \citep{hubisz2020inference}. A central quantity
in all of these sampling methods is the conditional sampling probability $p(G |
D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization of
genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a
frequentist regularizer for the ARG $G$.

The likelihood $p(D | G)$ is effectively always the distribution of a Poisson
process on the ancestral material along the edges of $G$ \citep[Eq.\
(2)]{mahmoudi2022bayesian}. Hence it is typically straightforward to evaluate
in principle, but choosing the right data structure can still have a dramatic
effect on the efficiency of evaluation \citep{mahmoudi2022bayesian}. Especially
in the Bayesian case, $p(G)$ is typically specified as the sampling
distribution of a stochastic process, such as the coalescent with
recombination. Two contemporary examples are \citet{mahmoudi2022bayesian} and
\citet{guo2022recombination}. Evaluating the distribution of the coalescent with
recombination (or related models) requires knowledge of the number of extant
lineages and number of links available for recombination (i.e.\ ones at which a
recombination would split ancestral material) at each time \citep[Eq.\
(3)]{mahmoudi2022bayesian}. These require a data structure from which shared
edges on different trees are easy to identify, and which encodes recombination
events and times explicitly, e.g.\ using nodes.
We will provide a more detailed discussion on sampling probability evaluation in the appendix.
\end{document}

>>>>

@@

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on specific details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
We show that eARGs are limited in the patterns
of genetic inheritance that can be represented,
and require a precise description of the details of all recombination events.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[][AND OTHERS]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is a lack of consensus on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks for the phrase Gideon and Alex!
If not addressed, these problems will
impede progress in applying genetic genealogies to understand the biological world.

@@

To make the following discussions precise, we require
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% This sentence needs another pass
A node in an oriented tree can have
zero, one (see the next section for discussion of ``unary'' nodes),
or many children.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree (or in any other form in which
the tree nodes are explicitly labelled with the identity of the
corresponding ARG node),
along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, under this
definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.
The internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes (see the next section) marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees (see section XXX).
If the local trees contain labelled internal nodes, precisely
describing the corresponding path through the graph for each
genome interval, then there is exactly as
much information in them as in an ARG.

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
% JK: Relate is inferring trees independently within segments,
% it's just using a dynamic algorithm to decide what those segments
% are, right? That's how it differs from things like Espalier which
% assume a set of input recomb breakpoints.
independently within genome segments,
and subsequently stitching them together into an ARG.
Two recent methods take this approach,
\relate~\citep{speidel2019method} and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among a small number of samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

@@

TODO mention \citep{schaefer2021ancestral} somewhere.

@@

An alternative approach is to treat the ARG as a latent parameter to be
averaged out by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, kuhner2006lamarc, nielsen2000estimation, wang2008bayesian,
wang2009population, fallon2013acg, mahmoudi2022bayesian}.
These methods operate on representations
of the ``little ARG'', and are extremely computationally expensive, being
applicable to at most hundreds of samples consisting of tens or hundreds of
kilobases with human-like parameters. State-of-the-art methods rely on cheaper,
approximate models \citep{didelot2010inference, heine2018bridging,
hubisz2020mapping,hubisz2020inference, medina2020speeding}. The most scalable
method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to
dozens of mammal-like genomes \citep{hubisz2020inference}.

Methods to sample ARGs generate a ``cloud'' of estimates
\citep{kuhner2017consensus} develop

% Not sure we want this content now? We're not really talking about
% evaluation of liklihoods any more.
% A central quantity
% in all of these sampling methods is the conditional sampling probability $p(G |
% D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization of
% genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
% of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a
% frequentist regularizer for the ARG $G$.

% The likelihood $p(D | G)$ is effectively always the distribution of a Poisson
% process on the ancestral material along the edges of $G$ \citep[Eq.\
% (2)]{mahmoudi2022bayesian}. Hence it is typically straightforward to evaluate
% in principle, but choosing the right data structure can still have a dramatic
% effect on the efficiency of evaluation \citep{mahmoudi2022bayesian}. Especially
% in the Bayesian case, $p(G)$ is typically specified as the sampling
% distribution of a stochastic process, such as the coalescent with
% recombination. Two contemporary examples are \citet{mahmoudi2022bayesian} and
% \citet{guo2022recombination}. Evaluating the distribution of the coalescent with
% recombination (or related models) requires knowledge of the number of extant
% lineages and number of links available for recombination (i.e.\ ones at which a
% recombination would split ancestral material) at each time \citep[Eq.\
% (3)]{mahmoudi2022bayesian}. These require a data structure from which shared
% edges on different trees are easy to identify, and which encodes recombination
% events and times explicitly, e.g.\ using nodes.

\end{document}
